---
title: "TabPFN vs XGBoost Benchmark"
output: rmarkdown::html_vignette
knitr:
  opts_chunk:
    dev: "pdf"
vignette: >
  %\VignetteIndexEntry{TabPFN vs XGBoost Benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("tabpfn") &&
         requireNamespace("parsnip",   quietly = TRUE) &&
         requireNamespace("workflows", quietly = TRUE) &&
         requireNamespace("recipes",   quietly = TRUE) &&
         requireNamespace("rsample",   quietly = TRUE) &&
         requireNamespace("yardstick", quietly = TRUE) &&
         requireNamespace("tune",      quietly = TRUE) &&
         requireNamespace("xgboost",   quietly = TRUE) &&
         requireNamespace("mlbench",   quietly = TRUE) &&
         requireNamespace("ggplot2",   quietly = TRUE) &&
         requireNamespace("tidyr",     quietly = TRUE)
)
```

This vignette compares **TabPFN** with **XGBoost** on small-to-medium tabular
datasets using 10-fold cross-validation through tidymodels. TabPFN is a
transformer-based model that performs in-context learning and typically excels
on datasets with fewer than a few thousand samples -- exactly the regime where
tree-based methods can struggle with limited data.

```{r libs, message = FALSE}
library(rtabpfn)
library(parsnip)
library(workflows)
library(recipes)
library(rsample)
library(yardstick)
library(tune)
library(mlbench)
library(ggplot2)
library(tidyr)
library(dplyr)
```

## Helper Function

A small wrapper that runs 10-fold CV for a given model spec and recipe,
then collects the metric summaries.

```{r helper}
run_cv <- function(spec, rec, folds, metrics) {
  wf <- workflow() |>
    add_recipe(rec) |>
    add_model(spec)
  res <- fit_resamples(wf, resamples = folds, metrics = metrics,
                       control = control_resamples(save_pred = FALSE))
  collect_metrics(res)
}
```

## Model Specifications

```{r specs}
tabpfn_clf <- tabpfn_model(
  mode         = "classification",
  n_estimators = 8L,
  device       = "cpu"
)

xgb_clf <- boost_tree(trees = 300, learn_rate = 0.1, tree_depth = 6) |>
  set_engine("xgboost") |>
  set_mode("classification")

tabpfn_reg <- tabpfn_model(
  mode         = "regression",
  n_estimators = 8L,
  device       = "cpu"
)

xgb_reg <- boost_tree(trees = 300, learn_rate = 0.1, tree_depth = 6) |>
  set_engine("xgboost") |>
  set_mode("regression")
```

---

# Classification Benchmarks

## 1. Ionosphere (351 x 34)

Radar returns from the ionosphere -- binary classification.

```{r clf-ionosphere}
data(Ionosphere)
iono <- Ionosphere[, -2]  # drop constant column V2

set.seed(42)
iono_folds <- vfold_cv(iono, v = 10, strata = Class)
iono_rec   <- recipe(Class ~ ., data = iono)
clf_metrics <- metric_set(accuracy, roc_auc)

iono_tabpfn <- run_cv(tabpfn_clf, iono_rec, iono_folds, clf_metrics) |>
  mutate(model = "TabPFN")
iono_xgb <- run_cv(xgb_clf, iono_rec, iono_folds, clf_metrics) |>
  mutate(model = "XGBoost")
iono_results <- bind_rows(iono_tabpfn, iono_xgb) |>
  mutate(dataset = "Ionosphere")
iono_results |> select(model, .metric, mean, std_err)
```

## 2. Pima Indians Diabetes (768 x 9)

Predicting diabetes onset -- binary classification.

```{r clf-pima}
data(PimaIndiansDiabetes)

set.seed(42)
pima_folds <- vfold_cv(PimaIndiansDiabetes, v = 10, strata = diabetes)
pima_rec   <- recipe(diabetes ~ ., data = PimaIndiansDiabetes)

pima_tabpfn <- run_cv(tabpfn_clf, pima_rec, pima_folds, clf_metrics) |>
  mutate(model = "TabPFN")
pima_xgb <- run_cv(xgb_clf, pima_rec, pima_folds, clf_metrics) |>
  mutate(model = "XGBoost")
pima_results <- bind_rows(pima_tabpfn, pima_xgb) |>
  mutate(dataset = "Pima")
pima_results |> select(model, .metric, mean, std_err)
```

## 3. Glass (214 x 10)

Identifying glass type from oxide content -- multi-class classification (6 classes).

```{r clf-glass}
data(Glass)

set.seed(42)
glass_folds <- vfold_cv(Glass, v = 10, strata = Type)
glass_rec   <- recipe(Type ~ ., data = Glass)

glass_tabpfn <- run_cv(tabpfn_clf, glass_rec, glass_folds, clf_metrics) |>
  mutate(model = "TabPFN")
glass_xgb <- run_cv(xgb_clf, glass_rec, glass_folds, clf_metrics) |>
  mutate(model = "XGBoost")
glass_results <- bind_rows(glass_tabpfn, glass_xgb) |>
  mutate(dataset = "Glass")
glass_results |> select(model, .metric, mean, std_err)
```

## Classification Summary

```{r clf-summary, fig.width = 8, fig.height = 4}
clf_all <- bind_rows(iono_results, pima_results, glass_results)

ggplot(clf_all, aes(x = dataset, y = mean, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    position = position_dodge(width = 0.7), width = 0.2
  ) +
  facet_wrap(~ .metric, scales = "free_y") +
  scale_fill_manual(values = c("TabPFN" = "#4E79A7", "XGBoost" = "#E15759")) +
  labs(x = NULL, y = "Mean (10-fold CV)", fill = NULL,
       title = "Classification: TabPFN vs XGBoost") +
  theme_minimal() +
  theme(legend.position = "top")
```

---

# Regression Benchmarks

## 1. Boston Housing (506 x 13)

Predicting median home value from neighbourhood characteristics.

```{r reg-boston}
data(BostonHousing)

set.seed(42)
boston_folds <- vfold_cv(BostonHousing, v = 10)
boston_rec   <- recipe(medv ~ ., data = BostonHousing)
reg_metrics <- metric_set(rmse, rsq)

boston_tabpfn <- run_cv(tabpfn_reg, boston_rec, boston_folds, reg_metrics) |>
  mutate(model = "TabPFN")
boston_xgb <- run_cv(xgb_reg, boston_rec, boston_folds, reg_metrics) |>
  mutate(model = "XGBoost")
boston_results <- bind_rows(boston_tabpfn, boston_xgb) |>
  mutate(dataset = "Boston")
boston_results |> select(model, .metric, mean, std_err)
```

## 2. Servo (167 x 5)

Predicting rise time from motor/screw configuration -- a very small dataset
where TabPFN's in-context learning should shine.

```{r reg-servo}
data(Servo)

set.seed(42)
servo_folds <- vfold_cv(Servo, v = 10)
servo_rec   <- recipe(Class ~ ., data = Servo)

servo_tabpfn <- run_cv(tabpfn_reg, servo_rec, servo_folds, reg_metrics) |>
  mutate(model = "TabPFN")
servo_xgb <- run_cv(xgb_reg, servo_rec, servo_folds, reg_metrics) |>
  mutate(model = "XGBoost")
servo_results <- bind_rows(servo_tabpfn, servo_xgb) |>
  mutate(dataset = "Servo")
servo_results |> select(model, .metric, mean, std_err)
```

## Regression Summary

```{r reg-summary, fig.width = 8, fig.height = 4}
reg_all <- bind_rows(boston_results, servo_results)

ggplot(reg_all, aes(x = dataset, y = mean, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    position = position_dodge(width = 0.7), width = 0.2
  ) +
  facet_wrap(~ .metric, scales = "free_y") +
  scale_fill_manual(values = c("TabPFN" = "#4E79A7", "XGBoost" = "#E15759")) +
  labs(x = NULL, y = "Mean (10-fold CV)", fill = NULL,
       title = "Regression: TabPFN vs XGBoost") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Combined Results Table

```{r results-table}
all_results <- bind_rows(clf_all, reg_all) |>
  select(dataset, model, .metric, mean, std_err) |>
  mutate(mean = round(mean, 4), std_err = round(std_err, 4))

all_results |>
  pivot_wider(
    names_from = model,
    values_from = c(mean, std_err),
    names_glue = "{model}_{.value}"
  ) |>
  arrange(dataset, .metric)
```

## Notes

- TabPFN uses default hyperparameters (`n_estimators = 8`); XGBoost uses a
  reasonable default (`trees = 300, learn_rate = 0.1, tree_depth = 6`).
  Neither model is tuned -- this is a fair out-of-the-box comparison.
- TabPFN is designed for datasets with fewer than ~10,000 samples and ~500
  features. On larger datasets, tree-based methods like XGBoost are typically
  more practical.
- All benchmarks use `device = "cpu"`. GPU acceleration can significantly
  speed up TabPFN inference.
