---
title: "TabPFN Embeddings"
output: rmarkdown::html_vignette
knitr:
  opts_chunk:
    dev: "pdf"
vignette: >
  %\VignetteIndexEntry{TabPFN Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("tabpfn") &&
         requireNamespace("ggplot2", quietly = TRUE) &&
         requireNamespace("Rtsne",   quietly = TRUE) &&
         requireNamespace("uwot",    quietly = TRUE)
)
```

TabPFN's transformer backbone produces rich internal representations for each
observation. The rtabpfn package exposes these representations through three
functions:

- `tabpfn_embeddings()` -- extract embeddings from a fitted model.
- `tabpfn_embeddings_cv()` -- cross-validated embeddings so every sample is
  embedded by a model that never saw it during training.
- `plot_embeddings()` -- reduce to 2-D with t-SNE or UMAP and visualise with
  ggplot2.

These embeddings are useful for exploratory analysis, clustering, or as
features for a downstream model.

```{r library}
library(rtabpfn)
```

## Vanilla Embeddings (Classification)

Fit a classifier on the iris data and extract embeddings for the test
observations. The `aggregate` argument controls how embeddings from multiple
ensemble members are combined.

```{r embed-clf}
clf <- tabpfn_classifier(
  x            = iris[, 1:4],
  y            = iris$Species,
  n_estimators = 4L,
  device       = "cpu"
)

emb_mean   <- tabpfn_embeddings(clf, iris[, 1:4], aggregate = "mean")
emb_concat <- tabpfn_embeddings(clf, iris[, 1:4], aggregate = "concat")
emb_raw    <- tabpfn_embeddings(clf, iris[, 1:4], aggregate = "none")

dim(emb_mean)    # (n_samples, embedding_dim)
dim(emb_concat)  # (n_samples, n_estimators * embedding_dim)
dim(emb_raw)     # (n_estimators, n_samples, embedding_dim)
```

## Vanilla Embeddings (Regression)

The same interface works for regressors.

```{r embed-reg}
reg <- tabpfn_regressor(
  x            = mtcars[, -1],
  y            = mtcars$mpg,
  n_estimators = 4L,
  device       = "cpu"
)

emb_reg <- tabpfn_embeddings(reg, mtcars[, -1])
dim(emb_reg)
```

## Cross-Validated Embeddings

`tabpfn_embeddings_cv()` fits a fresh model on each fold and embeds only the
held-out samples. This avoids information leakage when the embeddings are used
as features for a second-stage model.

### Auto-generated folds

```{r cv-auto}
set.seed(42)
emb_cv <- tabpfn_embeddings_cv(
  x            = iris[, 1:4],
  y            = iris$Species,
  n_folds      = 5L,
  n_estimators = 4L,
  device       = "cpu"
)
dim(emb_cv)
```

### rsample folds

You can pass an `rsample` rset object for precise control over the splits.

```{r cv-rsample}
library(rsample)

folds <- vfold_cv(iris, v = 5)
emb_cv2 <- tabpfn_embeddings_cv(
  x            = iris[, 1:4],
  y            = iris$Species,
  folds        = folds,
  n_estimators = 4L,
  device       = "cpu"
)
dim(emb_cv2)
```

## Visualising Embeddings

`plot_embeddings()` reduces the embedding matrix to two dimensions with t-SNE
or UMAP and returns a ggplot2 scatter plot. Both methods are shown side by side
below.

```{r viz, fig.width = 6, fig.height = 4}
p_tsne <- plot_embeddings(emb_cv, labels = iris$Species, method = "tsne")
p_umap <- plot_embeddings(emb_cv, labels = iris$Species, method = "umap")

p_tsne
p_umap
```

## Extracting Coordinates

Pass `return_coords = TRUE` to get both the plot and the underlying 2-D
coordinates as a tibble.

```{r coords}
res <- plot_embeddings(
  emb_cv,
  labels        = iris$Species,
  method        = "umap",
  return_coords = TRUE
)

head(res$coords)
res$plot
```

## From a Parsnip Workflow

If you are using the tidymodels framework, `extract_embeddings()` can pull
embeddings directly from a fitted parsnip model or workflow.

```{r workflow}
library(parsnip)
library(workflows)
library(recipes)

spec <- tabpfn_model(mode = "classification", n_estimators = 4L, device = "cpu")
rec  <- recipe(Species ~ ., data = iris)
wf   <- workflow() |> add_recipe(rec) |> add_model(spec)

wf_fit <- fit(wf, data = iris)
emb_wf <- extract_embeddings(wf_fit, iris[1:10, 1:4])
dim(emb_wf)
```
