---
title: "Getting Started with rtabpfn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rtabpfn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("tabpfn")
)
```

rtabpfn provides an R interface to [TabPFN](https://github.com/automl/tabpfn), a transformer-based model for tabular data that performs in-context learning — it is trained once on synthetic data and then applies to any new dataset without gradient-based fine-tuning.

## Installation

Install the R package from GitHub:

```r
# install.packages("pak")
pak::pak("your-org/rtabpfn")
```

Then install the TabPFN Python package into a dedicated virtual environment:

```r
library(rtabpfn)
install_tabpfn()
```

`install_tabpfn()` creates a `"r-tabpfn"` virtual environment and installs the `tabpfn` Python package via pip. You only need to do this once.

## Classification

Use `tabpfn_classifier()` to train a classifier. The function accepts a data frame of predictors `x` and a factor (or character) vector of labels `y`.

```{r classification-fit}
library(rtabpfn)

# Prepare data
x_train <- iris[1:120, 1:4]
y_train <- iris$Species[1:120]
x_test  <- iris[121:150, 1:4]
y_test  <- iris$Species[121:150]

# Fit the classifier
clf <- tabpfn_classifier(
  x           = x_train,
  y           = y_train,
  n_estimators = 2L,
  device       = "cpu"
)
clf
```

Predict class labels with `type = "class"`:

```{r classification-class}
preds_class <- predict(clf, x_test, type = "class")
head(preds_class)

# Accuracy
mean(preds_class$.pred_class == y_test)
```

Predict class probabilities with `type = "prob"`:

```{r classification-prob}
preds_prob <- predict(clf, x_test, type = "prob")
head(preds_prob)
```

The probability tibble has one column per class, named `.pred_<classname>`.

## Regression

Use `tabpfn_regressor()` to train a regressor. The target `y` must be a numeric vector.

```{r regression-fit}
x_train_r <- mtcars[1:25, -1]
y_train_r  <- mtcars$mpg[1:25]
x_test_r   <- mtcars[26:32, -1]
y_test_r   <- mtcars$mpg[26:32]

reg <- tabpfn_regressor(
  x            = x_train_r,
  y            = y_train_r,
  n_estimators = 2L,
  device       = "cpu"
)
reg
```

Predict the mean with `type = "numeric"` (the default):

```{r regression-numeric}
preds_num <- predict(reg, x_test_r, type = "numeric")
head(preds_num)

# RMSE
sqrt(mean((preds_num$.pred - y_test_r)^2))
```

Predict quantiles with `type = "quantile"`. The result is a long tibble with columns `.row`, `.quantile`, and `.pred`:

```{r regression-quantile}
preds_q <- predict(reg, x_test_r,
                   type      = "quantile",
                   quantiles = c(0.1, 0.5, 0.9))
head(preds_q, 9)
```

## Model Versions

TabPFN ships several model checkpoints. Use the `model_version` parameter to select one:

| Value    | Description                 |
|----------|-----------------------------|
| `NULL`   | Default (v2.6)              |
| `"v2"`   | Original TabPFN v2          |
| `"v2.5"` | Intermediate release        |
| `"v2.6"` | Latest release (same as `NULL`) |

```{r model-version, eval = FALSE}
# Use the v2.5 checkpoint
clf_v25 <- tabpfn_classifier(
  x             = x_train,
  y             = y_train,
  model_version = "v2.5",
  n_estimators  = 2L,
  device        = "cpu"
)
```

## Key Hyperparameters

Both `tabpfn_classifier()` and `tabpfn_regressor()` share the following hyperparameters:

**`n_estimators`** (integer, default `8L`)

Number of ensemble members. More estimators improve robustness at the cost of speed. For quick experiments, `2L` or `4L` work well.

```{r n-estimators, eval = FALSE}
clf <- tabpfn_classifier(x_train, y_train, n_estimators = 4L, device = "cpu")
```

**`softmax_temperature`** (double, default `0.9`)

Controls the sharpness of the predictive distribution. Values below 1 make predictions more confident; values above 1 make them softer.

```{r softmax-temperature, eval = FALSE}
clf <- tabpfn_classifier(x_train, y_train, softmax_temperature = 0.7, device = "cpu")
```

**`device`** (character, default `"auto"`)

Where to run computations. Options are `"auto"` (GPU if available, else CPU), `"cpu"`, and `"cuda"`.

```{r device, eval = FALSE}
# Force CPU
clf_cpu <- tabpfn_classifier(x_train, y_train, device = "cpu")

# Force GPU (requires CUDA)
clf_gpu <- tabpfn_classifier(x_train, y_train, device = "cuda")
```

**`balance_probabilities`** (classifier only, logical, default `FALSE`)

When `TRUE`, the classifier adjusts predicted probabilities to account for class imbalance.

```{r balance, eval = FALSE}
clf_bal <- tabpfn_classifier(x_train, y_train, balance_probabilities = TRUE, device = "cpu")
```
