---
title: "TabPFN Demo: R Translation of the Python Hands-On Notebook"
output: rmarkdown::html_vignette
knitr:
  opts_chunk:
    dev: "pdf"
vignette: >
  %\VignetteIndexEntry{TabPFN Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
tabpfn_ok <- reticulate::py_module_available("tabpfn") &&
  requireNamespace("mlbench",  quietly = TRUE) &&
  requireNamespace("ranger",   quietly = TRUE) &&
  requireNamespace("ggplot2",  quietly = TRUE) &&
  requireNamespace("tidyr",    quietly = TRUE) &&
  requireNamespace("yardstick", quietly = TRUE)

knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  dev       = "pdf",
  fig.width = 7,
  fig.height = 5,
  eval      = tabpfn_ok
)
```

This vignette is a direct R translation of the
[TabPFN Hands-On Demo Python notebook](https://github.com/PriorLabs/TabPFN/blob/main/TabPFN_Demo_Local.ipynb),
covering the same topics in order: classification, regression, decision
boundaries, curve fitting, and uncertainty quantification  -  all using the
`rtabpfn` package and standard tidyverse tools.

```{r pkgs, message = FALSE}
library(rtabpfn)
library(dplyr)
library(tibble)
library(ggplot2)
library(rsample)
library(yardstick)
library(ranger)
library(mlbench)
library(tidyr)
```

---

## 1  Classification: Parkinson's Disease Detection

The notebook uses the UCI Parkinson's Disease dataset: 195 voice measurements
from 31 people (23 with Parkinson's). The task is binary classification
evaluated with ROC AUC.

```{r pk-load}
pk_url <- paste0(
  "https://archive.ics.uci.edu/ml/machine-learning-databases/",
  "parkinsons/parkinsons.data"
)
pk  <- read.csv(pk_url)
X_pk <- pk |> select(-name, -status)
y_pk <- factor(pk$status, levels = c(0, 1), labels = c("healthy", "parkinsons"))
cat("Dimensions:", nrow(X_pk), "x", ncol(X_pk), "\nClass balance:", table(y_pk), "\n")
```

### Quick train / test split

```{r pk-split}
set.seed(42)
split   <- initial_split(bind_cols(X_pk, status = y_pk), prop = 0.8,
                         strata = status)
pk_tr   <- training(split);  pk_te <- testing(split)
X_tr    <- pk_tr |> select(-status); y_tr <- pk_tr$status
X_te    <- pk_te |> select(-status); y_te <- pk_te$status
```

```{r pk-fit}
clf   <- tabpfn_classifier(X_tr, y_tr, n_estimators = 8L, device = "cpu")
probs <- predict(clf, X_te, type = "prob")

auc <- roc_auc(
  bind_cols(truth = y_te, probs),
  truth = truth, .pred_parkinsons, event_level = "second"
)
cat("TabPFN ROC AUC:", round(auc$.estimate, 4), "\n")
```

### 10-fold cross-validation comparison

```{r pk-cv, message = FALSE}
pk_df <- bind_cols(X_pk, status = y_pk)
folds <- vfold_cv(pk_df, v = 10, strata = status)

cv_clf <- lapply(folds$splits, function(spl) {
  tr  <- training(spl);  te <- testing(spl)
  Xtr <- tr |> select(-status); ytr <- tr$status
  Xte <- te |> select(-status); yte <- te$status

  # TabPFN
  t_clf <- tabpfn_classifier(Xtr, ytr, n_estimators = 4L, device = "cpu")
  p_t   <- predict(t_clf, Xte, type = "prob")
  auc_t <- roc_auc(bind_cols(truth = yte, p_t), truth, .pred_parkinsons,
                   event_level = "second")$.estimate

  # Random Forest (ranger)
  rf_fit <- ranger(status ~ ., data = tr, probability = TRUE, seed = 42,
                   verbose = FALSE)
  p_rf   <- as_tibble(predict(rf_fit, te)$predictions)
  auc_rf <- roc_auc(bind_cols(truth = yte, p_rf), truth, parkinsons,
                    event_level = "second")$.estimate

  tibble(TabPFN = auc_t, `Random Forest` = auc_rf)
})

cv_clf_df <- bind_rows(cv_clf) |>
  summarise(across(everything(), mean)) |>
  pivot_longer(everything(), names_to = "Model", values_to = "ROC AUC")

ggplot(cv_clf_df, aes(x = reorder(Model, `ROC AUC`), y = `ROC AUC`, fill = Model)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = round(`ROC AUC`, 3)), vjust = -0.4, size = 3.5) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(
    title    = "Parkinson's classification  -  10-fold CV ROC AUC",
    subtitle = "Higher is better",
    x = NULL, y = "ROC AUC"
  ) +
  theme_minimal(base_size = 12)
```

---

## 2  Regression: Boston Housing

Predict median house values (`medv`) from 13 socio-economic features.
The dataset mirrors the one used in the Python notebook.

```{r bos-load}
data(BostonHousing)
X_bos <- BostonHousing |>
  select(-medv) |>
  mutate(across(where(is.factor), as.numeric))
y_bos <- BostonHousing$medv
cat("Dimensions:", nrow(X_bos), "x", ncol(X_bos), "\n")
```

### Quick train / test split (67 / 33)

```{r bos-fit}
set.seed(42)
idx  <- sample(nrow(X_bos), floor(0.67 * nrow(X_bos)))
Xtr  <- X_bos[idx,  ]; ytr <- y_bos[idx]
Xte  <- X_bos[-idx, ]; yte <- y_bos[-idx]

reg      <- tabpfn_regressor(Xtr, ytr, n_estimators = 8L, device = "cpu")
preds    <- predict(reg, Xte)

cat(sprintf(
  "TabPFN  RMSE: %.4f   R2: %.4f\n",
  rmse_vec(yte, preds$.pred),
  cor(yte, preds$.pred)^2
))
```

### 10-fold cross-validation comparison

```{r bos-cv, message = FALSE}
bos_df    <- bind_cols(X_bos, medv = y_bos)
folds_bos <- vfold_cv(bos_df, v = 10)

cv_reg <- lapply(folds_bos$splits, function(spl) {
  tr   <- training(spl);  te <- testing(spl)
  Xtr_ <- tr |> select(-medv); ytr_ <- tr$medv
  Xte_ <- te |> select(-medv); yte_ <- te$medv

  # TabPFN
  rg     <- tabpfn_regressor(Xtr_, ytr_, n_estimators = 4L, device = "cpu")
  p_t    <- predict(rg, Xte_)$.pred
  rsq_t  <- cor(yte_, p_t)^2

  # Random Forest
  rf_fit  <- ranger(medv ~ ., data = tr, seed = 42, verbose = FALSE)
  p_rf    <- predict(rf_fit, te)$predictions
  rsq_rf  <- cor(yte_, p_rf)^2

  tibble(TabPFN = rsq_t, `Random Forest` = rsq_rf)
})

cv_reg_df <- bind_rows(cv_reg) |>
  summarise(across(everything(), mean)) |>
  pivot_longer(everything(), names_to = "Model", values_to = "R2")

ggplot(cv_reg_df, aes(x = reorder(Model, R2), y = R2, fill = Model)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = round(R2, 3)), vjust = -0.4, size = 3.5) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(
    title    = "Boston Housing regression  -  10-fold CV R2",
    subtitle = "Variance explained  -  higher is better",
    x = NULL, y = "R2"
  ) +
  theme_minimal(base_size = 12)
```

---

## 3  Decision Boundaries: Concentric Circles

Two interleaved classes arranged in concentric rings expose non-linear
behaviour in classifiers  -  the same toy problem used in the Python notebook.

```{r circles-data}
set.seed(42)

gen_ring <- function(n, r, noise = 0.1) {
  ang <- runif(n, 0, 2 * pi)
  tibble(x1 = r * cos(ang) + rnorm(n, 0, noise),
         x2 = r * sin(ang) + rnorm(n, 0, noise))
}

circles <- bind_rows(
  bind_rows(gen_ring(50, 1), gen_ring(100, 2), gen_ring(200, 4)) |>
    mutate(class = "A"),
  bind_rows(gen_ring(50, 1.3), gen_ring(100, 2.3), gen_ring(200, 4.3)) |>
    mutate(class = "B")
) |> mutate(class = factor(class))
```

```{r circles-fit}
clf_circ <- tabpfn_classifier(circles |> select(x1, x2), circles$class,
                               n_estimators = 4L, device = "cpu")

rf_circ  <- ranger(class ~ x1 + x2, data = circles, probability = TRUE,
                   seed = 42, verbose = FALSE)

# Prediction grid
grid <- expand.grid(
  x1 = seq(min(circles$x1) - 0.5, max(circles$x1) + 0.5, length.out = 80),
  x2 = seq(min(circles$x2) - 0.5, max(circles$x2) + 0.5, length.out = 80)
)
grid$tabpfn <- predict(clf_circ, grid, type = "prob")$.pred_B
grid$rf     <- predict(rf_circ,  grid)$predictions[, "B"]

grid_long <- grid |>
  pivot_longer(c(tabpfn, rf), names_to = "model", values_to = "prob_B") |>
  mutate(model = recode(model, tabpfn = "TabPFN", rf = "Random Forest"))

circles_long <- bind_rows(
  circles |> mutate(model = "TabPFN"),
  circles |> mutate(model = "Random Forest")
)

ggplot() +
  geom_raster(data = grid_long, aes(x1, x2, fill = prob_B), alpha = 0.65) +
  geom_point(data = circles_long, aes(x1, x2, colour = class), size = 0.5) +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#4575b4",
                       midpoint = 0.5, name = "P(B)") +
  scale_colour_manual(values = c(A = "#d73027", B = "#4575b4")) +
  facet_wrap(~model) +
  labs(title = "Decision boundaries  -  concentric circles",
       x = "x1", y = "x2", colour = "Class") +
  coord_equal() +
  theme_minimal(base_size = 11)
```

---

## 4  Regression Curve Fitting: sin(x) + x/10

A smooth 1-D function sampled at 40 points. The goal is to compare how well
each model captures the underlying shape, especially outside the training
distribution's density.

```{r sin-data}
set.seed(42)
N        <- 40
x_train  <- matrix(seq(0, 20, length.out = N), ncol = 1)
y_train  <- sin(x_train) + x_train / 10
x_test   <- matrix(seq(0, 20, length.out = 200), ncol = 1)
```

```{r sin-fit}
reg_sin <- tabpfn_regressor(as.data.frame(x_train), as.vector(y_train),
                             n_estimators = 8L, device = "cpu")
rf_sin  <- ranger(y ~ x,
                  data = tibble(x = x_train[, 1], y = y_train[, 1]),
                  seed = 42, verbose = FALSE)

pred_t  <- predict(reg_sin, as.data.frame(x_test))$.pred
pred_rf <- predict(rf_sin, tibble(x = x_test[, 1]))$predictions
y_true  <- sin(x_test) + x_test / 10   # ground truth on test grid

curves <- tibble(
  x = rep(x_test[, 1], 3),
  y = c(pred_t, pred_rf, y_true[, 1]),
  source = rep(c("TabPFN", "Random Forest", "True function"), each = 200)
)
train_pts <- tibble(x = x_train[, 1], y = y_train[, 1])

ggplot() +
  geom_line(data = curves,
            aes(x, y, colour = source, linewidth = source, linetype = source)) +
  geom_point(data = train_pts, aes(x, y), colour = "black", size = 2) +
  scale_colour_manual(
    values = c("TabPFN" = "#4575b4", "Random Forest" = "#d73027",
               "True function" = "grey40")
  ) +
  scale_linewidth_manual(
    values = c("TabPFN" = 1, "Random Forest" = 1, "True function" = 0.6)
  ) +
  scale_linetype_manual(
    values = c("TabPFN" = "solid", "Random Forest" = "solid",
               "True function" = "dashed")
  ) +
  labs(
    title    = "1-D regression: sin(x) + x/10",
    subtitle = "Black points = training data",
    x = "x", y = "y",
    colour = NULL, linewidth = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 5  Uncertainty Quantification

TabPFN natively estimates the full predictive distribution. We fit it to
data with heteroscedastic (growing) noise and plot quantile prediction
intervals  -  the same demonstration from the Python notebook's final section.

```{r het-data}
set.seed(42)
n_pts   <- 200
x_het   <- matrix(seq(0, 100, length.out = n_pts), ncol = 1)
y_true_het <- 0.8 * x_het[, 1]
y_noisy    <- y_true_het + rnorm(n_pts, 0, 0.1 * x_het[, 1])

x_test_het <- matrix(seq(0, 100, length.out = 120), ncol = 1)
```

```{r het-fit}
reg_het <- tabpfn_regressor(as.data.frame(x_het), y_noisy,
                             n_estimators = 4L, device = "cpu")

q_levels <- c(0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95)
q_preds  <- predict(reg_het, as.data.frame(x_test_het),
                    type = "quantile", quantiles = q_levels)

# Pivot to wide format for ribbon plotting
q_wide <- q_preds |>
  mutate(q_label = paste0("q", sprintf("%03d", as.integer(.quantile * 100)))) |>
  pivot_wider(id_cols = .row, names_from = q_label, values_from = .pred) |>
  mutate(x = x_test_het[.row, 1])

ggplot(q_wide, aes(x = x)) +
  geom_ribbon(aes(ymin = q005, ymax = q095), fill = "#4575b4", alpha = 0.15) +
  geom_ribbon(aes(ymin = q010, ymax = q090), fill = "#4575b4", alpha = 0.20) +
  geom_ribbon(aes(ymin = q025, ymax = q075), fill = "#4575b4", alpha = 0.30) +
  geom_line(aes(y = q050), colour = "#4575b4", linewidth = 1, linetype = "dashed") +
  geom_line(data = tibble(x = x_het[, 1], y = y_true_het),
            aes(x = x, y = y), colour = "black", linewidth = 0.8) +
  geom_point(data = tibble(x = x_het[, 1], y = y_noisy),
             aes(x = x, y = y), colour = "#d73027", size = 0.8, alpha = 0.4) +
  labs(
    title    = "TabPFN quantile regression  -  heteroscedastic noise",
    subtitle = paste(
      "Ribbons: 5-95%, 10-90%, 25-75% intervals",
      "| dashed = median | black = true line | red = training data"
    ),
    x = "x", y = "y"
  ) +
  theme_minimal(base_size = 12)
```

The widening intervals correctly capture the increasing variance: for small
`x` the model is confident, while for large `x` the uncertainty grows
proportionally to the noise scale  -  exactly the heteroscedastic structure
injected into the data.
