---
title: "TabPFN with tidymodels"
output: rmarkdown::html_vignette
knitr:
  opts_chunk:
    dev: "pdf"
vignette: >
  %\VignetteIndexEntry{TabPFN with tidymodels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = reticulate::py_module_available("tabpfn") &&
         requireNamespace("parsnip",   quietly = TRUE) &&
         requireNamespace("workflows", quietly = TRUE) &&
         requireNamespace("recipes",   quietly = TRUE)
)
```

rtabpfn integrates with the [tidymodels](https://www.tidymodels.org/) ecosystem through a custom parsnip engine. This lets you use TabPFN inside the same `recipe` → `workflow` → `fit` → `predict` pipeline you already know.

## Overview

`tabpfn_model()` creates a parsnip model specification. It accepts:

- `mode`: `"classification"` or `"regression"`
- `n_estimators`: number of ensemble members (default `8L`)
- `softmax_temperature`: predictive sharpness (default `0.9`)
- `device`: `"auto"`, `"cpu"`, or `"cuda"` (default `"auto"`)
- `model_version`: TabPFN checkpoint — `NULL` (default, v2.6), `"v2"`, `"v2.5"`, or `"v2.6"`

All hyperparameters can be wrapped in `tune::tune()` for grid search.

## Classification Workflow

```{r libs-class}
library(rtabpfn)
library(parsnip)
library(workflows)
library(recipes)
library(yardstick)
library(rsample)
```

Split the iris data, build a recipe, and fit a workflow:

```{r classification-split}
set.seed(42)
iris_split <- initial_split(iris, prop = 0.8, strata = Species)
iris_train <- training(iris_split)
iris_test  <- testing(iris_split)
```

```{r classification-recipe}
iris_rec <- recipe(Species ~ ., data = iris_train) |>
  step_normalize(all_numeric_predictors())
```

```{r classification-spec}
clf_spec <- tabpfn_model(
  mode          = "classification",
  n_estimators  = 2L,
  device        = "cuda"
)
```

```{r classification-workflow}
clf_wf <- workflow() |>
  add_recipe(iris_rec) |>
  add_model(clf_spec)

clf_fit <- fit(clf_wf, data = iris_train)
clf_fit
```

Predict class labels and evaluate accuracy:

```{r classification-predict}
preds_class <- predict(clf_fit, iris_test, type = "class")
preds_class

results_class <- iris_test |>
  dplyr::select(Species) |>
  dplyr::bind_cols(preds_class)

accuracy(results_class, truth = Species, estimate = .pred_class)
```

Predict class probabilities:

```{r classification-prob}
preds_prob <- predict(clf_fit, iris_test, type = "prob")
head(preds_prob)
```

## Regression Workflow

```{r libs-reg}
library(yardstick)
```

```{r regression-split}
set.seed(42)
mtcars_split <- initial_split(mtcars, prop = 0.8)
mtcars_train <- training(mtcars_split)
mtcars_test  <- testing(mtcars_split)
```

```{r regression-recipe}
mtcars_rec <- recipe(mpg ~ ., data = mtcars_train) |>
  step_normalize(all_numeric_predictors())
```

```{r regression-spec}
reg_spec <- tabpfn_model(
  mode          = "regression",
  n_estimators  = 2L,
  device        = "cpu"
)
```

```{r regression-workflow}
reg_wf <- workflow() |>
  add_recipe(mtcars_rec) |>
  add_model(reg_spec)

reg_fit <- fit(reg_wf, data = mtcars_train)
reg_fit
```

Predict numeric outcomes and compute RMSE:

```{r regression-predict}
preds_num <- predict(reg_fit, mtcars_test, type = "numeric")
head(preds_num)

results_reg <- mtcars_test |>
  dplyr::select(mpg) |>
  dplyr::bind_cols(preds_num)

rmse(results_reg, truth = mpg, estimate = .pred)
```

## Cross-Validation

Use `rsample::vfold_cv()` with `tune::fit_resamples()` to estimate out-of-sample performance without hyperparameter tuning:

```{r cv-setup}
library(tune)
```

```{r cv-run}
set.seed(123)
iris_folds <- vfold_cv(iris_train, v = 5, strata = Species)

cv_spec <- tabpfn_model(
  mode          = "classification",
  n_estimators  = 2L,
  device        = "cpu"
)

cv_wf <- workflow() |>
  add_recipe(iris_rec) |>
  add_model(cv_spec)

cv_results <- fit_resamples(
  cv_wf,
  resamples = iris_folds,
  metrics   = metric_set(accuracy, roc_auc),
  control   = control_resamples(save_pred = TRUE)
)

collect_metrics(cv_results)
```

## Hyperparameter Tuning

Wrap hyperparameters with `tune()` to search over a grid. This is illustrative — running a full grid search can be slow, so this chunk is not evaluated.

```{r tuning, eval = FALSE}
library(tune)
library(dials)

set.seed(42)
iris_folds_tune <- vfold_cv(iris_train, v = 3, strata = Species)

tune_spec <- tabpfn_model(
  mode                = "classification",
  n_estimators        = tune(),
  softmax_temperature = tune(),
  device              = "cpu"
)

tune_rec <- recipe(Species ~ ., data = iris_train) |>
  step_normalize(all_numeric_predictors())

tune_wf <- workflow() |>
  add_recipe(tune_rec) |>
  add_model(tune_spec)

# Define a manual grid
tune_grid <- expand.grid(
  n_estimators        = c(4L, 8L, 16L),
  softmax_temperature = c(0.7, 0.9, 1.1)
)

tune_results <- tune_grid(
  tune_wf,
  resamples = iris_folds_tune,
  grid      = tune_grid,
  metrics   = metric_set(accuracy)
)

# Show best configurations
show_best(tune_results, metric = "accuracy")

# Finalise the workflow with the best hyperparameters
best_params <- select_best(tune_results, metric = "accuracy")
final_wf    <- finalize_workflow(tune_wf, best_params)
final_fit   <- fit(final_wf, data = iris_train)
```
